---
title: "The Censoring Framework for k-out-of-n Systems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Censoring Framework for k-out-of-n Systems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

## The fundamental problem

Suppose you observe system-level failure times from a multi-component system.
The system is a "black box": you record *when* it fails, but not which
components failed, nor in what order. Can you still learn about the individual
component lifetime distributions?

This is the core question that `kofn` addresses. The answer depends on the
system structure, and the key insight is that *system structure induces
structured censoring on component lifetimes*. This vignette develops that
insight for $k$-out-of-$n$ systems.

```{r}
library(kofn)
```


## Series vs parallel: two kinds of incomplete data

A **series system** ($k = n$) fails when its *first* component fails:
$T = \min(T_1, \ldots, T_m)$. At system failure, one component has just
failed and all others are still working. The incomplete-data problem is:
*which component caused the failure?* This is the classical **masked
cause-of-failure** problem.

A **parallel system** ($k = 1$) fails when its *last* component fails:
$T = \max(T_1, \ldots, T_m)$. At system failure, all components have already
failed. The incomplete-data problem is different: *when did each component
fail?* This is a **censoring** problem --- we know every component failed
before $T$, but we do not know the individual failure times.

```{r}
# Create both system types
series_sys <- series_system(3)
parallel_sys <- parallel_system(3)

# Same component lifetimes, different system behavior
comp_times <- c(1.5, 3.2, 2.8)

system_lifetime(series_sys, comp_times)   # min = 1.5
system_lifetime(parallel_sys, comp_times) # max = 3.2
```

The `system_censoring()` function reveals what each system structure tells us
about the individual components:

```{r}
# Series: component 1 observed exactly, components 2 and 3 right-censored
system_censoring(series_sys, comp_times)

# Parallel: component 2 observed exactly, components 1 and 3 left-censored
system_censoring(parallel_sys, comp_times)
```

In the series case, surviving components have $T_j > T$ (they outlasted the
system), so they are **right-censored**. In the parallel case, already-failed
components have $T_j < T$ (they failed before the system did), so they are
**left-censored**. The censoring direction reverses between the two extremes.


## The censoring equivalence for k-out-of-n systems

A $k$-out-of-$n$ system functions as long as at least $k$ of its $n$
components survive. It fails when the $(n - k + 1)$-th component fails, so
the system lifetime is the $(n - k + 1)$-th order statistic:
$$T = T_{(n-k+1)}.$$

At system failure time $T = t$, the components partition into three groups:

1. **One component observed exactly** --- the $(n-k+1)$-th to fail, whose
   lifetime equals $T$.
2. **$n - k$ components left-censored** at $T$ --- they failed before the
   system ($T_j < T$), but we do not know when.
3. **$k - 1$ components right-censored** at $T$ --- they are still functioning
   ($T_j > T$), but we do not know how long they will last.
4. **All identities are latent** --- we do not know which components belong to
   which group.

This is the **censoring equivalence**: observing a $k$-out-of-$n$ system
failure is informationally equivalent to observing 1 exact, $n - k$
left-censored, and $k - 1$ right-censored component lifetimes with unknown
assignment.

```{r}
# 2-out-of-4 system: k=2, n=4
sys_2of4 <- kofn_system(2, 4)
comp_times_4 <- c(1.5, 3.2, 2.8, 4.1)

cens <- system_censoring(sys_2of4, comp_times_4)
cens
```

```{r}
# Count the censoring types
table(cens$status)
```

For this 2-out-of-4 system ($k = 2$, $n = 4$), we expect:

- $n - k = 2$ left-censored (failed before the system)
- $k - 1 = 1$ right-censored (still functioning)
- 1 exact (the critical component)


## Censoring patterns across k

The parameter $k$ controls the split between left- and right-censored
components. As $k$ increases from 1 (parallel) to $n$ (series), right-censored
components replace left-censored ones:

| System | $k$ | Left-censored | Exact | Right-censored |
|--------|-----|:-------------:|:-----:|:--------------:|
| Parallel ($k=1$)  | 1 | $n-1$ | 1 | 0 |
| 2-out-of-$n$      | 2 | $n-2$ | 1 | 1 |
| ...                | ... | ... | 1 | ... |
| $(n\!-\!1)$-out-of-$n$ | $n-1$ | 1 | 1 | $n-2$ |
| Series ($k=n$)    | $n$ | 0 | 1 | $n-1$ |

The total number of censored components is always $n - 1$; only the
*direction* changes. Let us verify this empirically by simulating many
system failures and tallying the censoring patterns.

```{r censoring-pattern-plot, fig.height = 5}
set.seed(42)
n_comp <- 5
n_sim <- 2000
rates <- c(0.5, 1.0, 1.5, 2.0, 2.5)

# For each k, simulate systems and count censoring types
results <- data.frame(
  k = integer(), left = numeric(), exact = numeric(), right = numeric()
)

for (k in 1:n_comp) {
  sys <- kofn_system(k, n_comp)
  left_count <- exact_count <- right_count <- 0
  for (i in seq_len(n_sim)) {
    times <- rexp(n_comp, rate = rates)
    cens <- system_censoring(sys, times)
    left_count  <- left_count  + sum(cens$status == "left")
    exact_count <- exact_count + sum(cens$status == "exact")
    right_count <- right_count + sum(cens$status == "right")
  }
  results <- rbind(results, data.frame(
    k = k,
    left  = left_count  / n_sim,
    exact = exact_count / n_sim,
    right = right_count / n_sim
  ))
}

# Plot stacked bar chart
cols <- c("left" = "#e41a1c", "exact" = "#4daf4a", "right" = "#377eb8")
bp <- barplot(
  t(as.matrix(results[, c("left", "exact", "right")])),
  beside = FALSE,
  names.arg = results$k,
  col = cols,
  xlab = "k (system parameter)",
  ylab = "Average component count per observation",
  main = "Censoring pattern by system type (n = 5)",
  ylim = c(0, n_comp + 0.5)
)
legend("topright",
  legend = c("Left-censored", "Exact", "Right-censored"),
  fill = cols, bty = "n"
)
```

As expected, each observation always yields exactly 1 exact component
lifetime and $n - 1 = 4$ censored ones. The $k$ parameter controls only
the left/right split.


## Parallel systems (k = 1): the deeply-censored fast components

The parallel system is the most challenging case: all $m - 1$ non-critical
components are left-censored. Which components are hardest to estimate?

A "fast" component (high failure rate $\lambda_j$) tends to fail early ---
long before the system fails. Its lifetime is deeply left-censored: we know
$T_j < T$, but $T_j$ may be far below $T$. A "slow" component (low
$\lambda_j$) tends to be the last to fail, so its lifetime is often observed
exactly.

The probability $P(T_j = T)$ --- that component $j$ determines the system
lifetime --- governs how much information each system observation carries
about $\lambda_j$.

```{r information-demo, fig.height = 5}
set.seed(123)
n_sim <- 5000
rates <- c(0.5, 1.0, 2.0, 4.0)
m <- length(rates)
sys <- parallel_system(m)

# Track which component is critical (last to fail)
critical_counts <- integer(m)
for (i in seq_len(n_sim)) {
  times <- rexp(m, rate = rates)
  cens <- system_censoring(sys, times)
  critical_counts[cens$critical] <- critical_counts[cens$critical] + 1L
}

# P(T_j = T) is the proportion of times component j is critical
p_critical <- critical_counts / n_sim

par(mfrow = c(1, 2), mar = c(5, 4, 3, 1))

# Left panel: rates
barplot(rates, names.arg = paste0("Comp ", 1:m),
        col = "#377eb8", ylab = expression(lambda[j]),
        main = "Component failure rates")

# Right panel: P(T_j = T)
barplot(p_critical, names.arg = paste0("Comp ", 1:m),
        col = "#e41a1c", ylab = expression(P(T[j] == T)),
        main = "Probability of being critical")
```

Component 1 (the slowest, $\lambda_1 = 0.5$) is critical in roughly
`r round(100 * p_critical[1])`% of observations. Component 4 (the fastest,
$\lambda_4 = 4$) is critical in only
`r round(100 * p_critical[4], 1)`% of observations. This asymmetry
translates directly into estimation precision: slow components are
well-estimated from system data, while fast components require enormous
sample sizes or additional monitoring.


## Why this matters: predicting estimation difficulty

The censoring framework does more than restate the likelihood --- it provides
a predictive tool. Before collecting data, one can compute $P(T_j = T)$ for
each component and anticipate which parameters will be hard to estimate.

For exponential components with rates $\lambda_1, \ldots, \lambda_m$:
$$P(T_j = T) = \int_0^\infty f_j(t) \prod_{i \neq j} F_i(t) \, dt$$
which in the `kofn` package is computed by the inclusion-exclusion expansion.
Components with $P(T_j = T) \approx 0$ are effectively invisible to
system-level data.

```{r estimation-demo}
# Generate data and fit a parallel system model
set.seed(99)
model <- kofn(k = 1, m = 3, family = "exponential")
true_rates <- c(0.5, 1.0, 3.0)

gen <- rdata(model)
df <- gen(theta = true_rates, n = 500)

fit_fn <- fit(model)
result <- fit_fn(df)

# Compare estimates to truth
est <- coef(result)
se <- sqrt(diag(vcov(result)))
cat("Component | True rate | Estimate |   SE   | Relative SE\n")
cat("---------+-----------+----------+--------+------------\n")
for (j in seq_along(true_rates)) {
  cat(sprintf("    %d     |   %.1f     |  %.3f   | %.3f  |   %.1f%%\n",
              j, true_rates[j], est[j], se[j], 100 * se[j] / est[j]))
}
```

The slow component ($\lambda_1 = 0.5$) is estimated precisely.
The fast component ($\lambda_3 = 3.0$) has a much larger relative standard
error --- exactly as the censoring framework predicts.


## Visualizing censoring depth

To see the censoring structure concretely, we can look at latent component
lifetimes alongside the observed system lifetime. The gap between a
component's true failure time and the system failure time shows how deeply
it is censored.

```{r censoring-depth-plot, fig.height = 5.5}
set.seed(7)
n_show <- 30
true_rates <- c(0.5, 1.5, 3.0)
m <- length(true_rates)
sys <- parallel_system(m)

# Generate data and extract latent lifetimes
model <- kofn(k = 1, m = m, family = "exponential")
gen <- rdata(model)
df <- gen(theta = true_rates, n = n_show)
comp_times <- attr(df, "comp_lifetimes")

par(mar = c(5, 4, 3, 1))
plot(NULL, xlim = c(0, max(comp_times) * 1.05), ylim = c(0.5, n_show + 0.5),
     xlab = "Time", ylab = "Observation",
     main = "Component lifetimes vs system lifetime (parallel, m = 3)")

comp_cols <- c("#377eb8", "#e41a1c", "#4daf4a")

for (i in seq_len(n_show)) {
  t_sys <- df$t[i]
  # System lifetime (vertical dashed line for this obs)
  segments(t_sys, i - 0.3, t_sys, i + 0.3, lwd = 2, col = "grey40")

  for (j in seq_len(m)) {
    y_offset <- (j - 2) * 0.15
    t_j <- comp_times[i, j]
    # Component failure time
    points(t_j, i + y_offset, pch = if (t_j == t_sys) 16 else 1,
           col = comp_cols[j], cex = 0.8)
    # Arrow from component failure to system failure (shows censoring depth)
    if (t_j < t_sys) {
      arrows(t_j, i + y_offset, t_sys, i + y_offset,
             length = 0.03, col = adjustcolor(comp_cols[j], alpha = 0.4))
    }
  }
}

legend("topright",
  legend = c(
    expression(lambda[1] == 0.5 ~ "(slow)"),
    expression(lambda[2] == 1.5 ~ "(medium)"),
    expression(lambda[3] == 3.0 ~ "(fast)")
  ),
  col = comp_cols, pch = 16, bty = "n", cex = 0.9
)
legend("right",
  legend = c("Exact (critical)", "Left-censored", "System failure"),
  pch = c(16, 1, NA), lty = c(NA, NA, 1), lwd = c(NA, NA, 2),
  col = c("black", "black", "grey40"), bty = "n", cex = 0.9
)
```

The slow component (blue, $\lambda_1 = 0.5$) is usually the last to fail
--- its failure time coincides with the system lifetime (filled circles). The
fast component (green, $\lambda_3 = 3.0$) fails early and is deeply
left-censored: the arrows show the gap between its true failure time (open
circle) and the system failure time (grey bar). This gap is the information
lost by observing only the system.


## Connection to the series case (maskedcauses)

When $k = n$, the $k$-out-of-$n$ system becomes a series system, and the
censoring equivalence reduces to the classical masked cause-of-failure
structure:

- 1 component observed exactly (the first to fail)
- $n - 1$ components right-censored (still functioning)
- The identity of the failed component may be masked

This is precisely the setting handled by the `maskedcauses` package. The
`kofn` package recovers this as a special case:

```{r series-equivalence}
# k-out-of-n with k = n is a series system
sys_series <- kofn_system(k = 4, n = 4)
comp_times <- c(2.1, 0.8, 3.5, 1.9)

cens_series <- system_censoring(sys_series, comp_times)
cens_series
```

Component 2 (the first to fail, at $t = 0.8$) is observed exactly, and all
others are right-censored. The unknown in the series case is *which*
component failed --- the classical masking problem. In the parallel case, the
unknown is *when* each component failed --- a censoring problem. The `kofn`
package treats both within a single framework.


## The censoring equivalence in code

The `system_censoring()` function works for any coherent system, not just
$k$-out-of-$n$ systems. Here is a bridge system example:

```{r bridge}
bridge <- bridge_system()
bridge

comp_times <- c(2.1, 1.3, 0.5, 3.8, 2.9)
system_censoring(bridge, comp_times)
```

For general coherent systems, the censoring pattern depends on the specific
realization of component lifetimes and the system's cut-set structure. The
$k$-out-of-$n$ case is special because the *counts* (how many left, exact,
right) are deterministic --- only the *identities* are random.


## Summary

The censoring framework provides a unified view of component estimation from
system data:

- **Every system observation** yields one exact component lifetime and $m - 1$
  censored ones. The system structure determines the censoring direction.

- **The parameter $k$** controls the left/right split: parallel ($k = 1$)
  gives all left-censoring, series ($k = n$) gives all right-censoring,
  intermediate $k$ gives a mixture.

- **Estimation difficulty** is governed by $P(T_j = T)$: components that
  rarely determine the system lifetime are deeply censored and hard to
  estimate.

- **The series case** ($k = n$) recovers the masked cause-of-failure problem
  handled by `maskedcauses`. The parallel case ($k = 1$) is the opposite
  extreme, where all components have failed but their individual failure
  times are left-censored.

This framework is developed formally as Propositions 1 and 5 in Towell (2026),
*"Component lifetime estimation from parallel system data: a censoring
approach."*
