---
title: "Exponential Parallel Systems: Closed-Form MLE via Inclusion-Exclusion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exponential Parallel Systems: Closed-Form MLE via Inclusion-Exclusion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>"
)

# Set to TRUE to regenerate long-running simulation results.
# With run_long = FALSE, only small demos run (< 30 seconds total).
run_long <- FALSE

library(kofn)
old_opts <- options(digits = 4)
```

# The Exponential Parallel System

A **parallel system** of $m$ independent components functions as long as at
least one component survives.  The system lifetime is
$$
T = \max(T_1, \ldots, T_m), \qquad T_j \sim \mathrm{Exp}(\lambda_j).
$$

The system density is the derivative of the system CDF
$F_{\mathrm{sys}}(t) = \prod_{j=1}^{m} F_j(t) = \prod_{j=1}^{m}(1 - e^{-\lambda_j t})$:
$$
f_{\mathrm{sys}}(t) = \sum_{j=1}^{m} f_j(t) \prod_{i \neq j} F_i(t)
  = \sum_{j=1}^{m} \lambda_j e^{-\lambda_j t} \prod_{i \neq j}
    \bigl(1 - e^{-\lambda_i t}\bigr).
$$

Each summand $w_j(t) = f_j(t)\prod_{i \neq j} F_i(t)$ is the contribution
from the event "component $j$ is the last to fail."

```{r model-setup}
# Create an exponential parallel system model with 3 components
model <- kofn(k = 1, m = 3, family = "exponential")
print(model)
```

The `kofn()` constructor with `k = 1` creates a parallel system.  The model
object carries the system structure, column name conventions, and dispatches
to methods for `loglik()`, `score()`, `hess_loglik()`, `fit()`, and `rdata()`.


# Inclusion-Exclusion Expansion

The product $\prod_{i}(1 - e^{-\lambda_i t})$ cannot be differentiated or
integrated term-by-term as written.  The **inclusion-exclusion expansion**
rewrites it as a finite signed sum of exponentials:
$$
\prod_{i=1}^{m} \bigl(1 - e^{-\lambda_i t}\bigr)
  = \sum_{S \subseteq \{1,\ldots,m\}} (-1)^{|S|}\, e^{-(\sum_{i \in S} \lambda_i)\, t}.
$$

This has $2^m$ terms --- practical for moderate $m$ (up to $\sim$15
components).  Because every term is a simple exponential $e^{-rt}$, all
integrals needed for censored-data likelihoods evaluate in closed form.

```{r ie-demo}
lam <- c(0.5, 0.3, 0.2)
ie <- ie_expand(lam)

# Display the expansion terms
data.frame(
  subset_size = sapply(seq_along(ie$sign), function(k) {
    sum(as.integer(intToBits(k - 1L))[1:3])
  }),
  sign = ie$sign,
  rate_sum = ie$rate_sum
)
```

The first row ($S = \emptyset$) has sign $+1$ and rate sum $0$, contributing
the constant term $1$.  The three singleton subsets contribute $-e^{-\lambda_j t}$,
the three pairs contribute $+e^{-(\lambda_i + \lambda_j)t}$, and the full set
contributes $-e^{-(\lambda_1 + \lambda_2 + \lambda_3)t}$.

We can verify the expansion numerically:

```{r ie-verify}
t_val <- 2.0

# Direct product
direct <- prod(1 - exp(-lam * t_val))

# Via IE expansion
ie_val <- sum(ie$sign * exp(-ie$rate_sum * t_val))

c(direct = direct, ie_expansion = ie_val, difference = direct - ie_val)
```

The key functions built on this expansion are:

- `w_j_exact(t, par, j)` --- evaluates $w_j(t)$ at a single time point
- `w_j_integral(a, b, par, j)` --- closed-form $\int_a^b w_j(t)\,dt$
- `F_sys_exp(t, par)` / `S_sys_exp(t, par)` --- system CDF and survival

```{r ie-functions}
# The three w_j contributions at t = 2
sapply(1:3, function(j) w_j_exact(t_val, lam, j))

# Their sum equals the system density
sum(sapply(1:3, function(j) w_j_exact(t_val, lam, j)))

# System CDF and survival at t = 2
c(F_sys = F_sys_exp(t_val, lam), S_sys = S_sys_exp(t_val, lam))
```


# Likelihood Under Different Observation Types

The IE expansion gives closed-form log-likelihood contributions for all four
observation types.  Let $\boldsymbol{\lambda} = (\lambda_1, \ldots, \lambda_m)$ and
$C_i \subseteq \{1, \ldots, m\}$ be the candidate set for observation $i$.

**Exact** observation at $t$:
$$
\ell_i = \log \sum_{j \in C_i} w_j(t;\, \boldsymbol{\lambda}).
$$

**Right-censored** at $t$ (system still functioning):
$$
\ell_i = \log S_{\mathrm{sys}}(t;\, \boldsymbol{\lambda}).
$$

**Left-censored** at $t$ (system already failed):
$$
\ell_i = \log \frac{\sum_{j \in C_i} \int_0^{t} w_j(s)\,ds}
{F_{\mathrm{sys}}(t;\, \boldsymbol{\lambda})}.
$$

**Interval-censored** on $(a, b]$:
$$
\ell_i = \log \frac{\sum_{j \in C_i} \int_a^{b} w_j(s)\,ds}
{F_{\mathrm{sys}}(b) - F_{\mathrm{sys}}(a)}.
$$

All integrals $\int_a^b w_j(s)\,ds$ are computed analytically by
`w_j_integral()`.  The `loglik(model)` closure handles all four types
automatically:

```{r loglik-demo}
set.seed(42)

# Generate data under Scheme 0 (system lifetime only, no censoring)
gen <- rdata(model)
df <- gen(theta = lam, n = 50, p = 1)
head(df, 5)

# Evaluate the log-likelihood at the true parameters
ll_fn <- loglik(model)
ll_fn(df, lam)

# Compare with a wrong parameter vector
ll_fn(df, c(1, 1, 1))
```


# Maximum Likelihood Estimation

## Data generation

We generate $n = 100$ system lifetimes from a 3-component parallel system
with rates $\boldsymbol{\lambda} = (0.5, 0.3, 0.2)$ under Scheme 0 (system
lifetime only --- no component-level information).

```{r generate-data}
set.seed(123)
theta_true <- c(0.5, 0.3, 0.2)
n <- 100

gen <- rdata(model)
df <- gen(theta = theta_true, n = n, p = 1)

cat("Observations:", nrow(df), "\n")
cat("Observation types:\n")
table(df$omega)
```

With `p = 1`, every non-critical component is included in the candidate set,
so all components are candidates for every observation.  This is the
Scheme 0 (black-box) setting: we observe only when the system fails, not
which component failed last.

## Fitting

The `fit(model)` closure performs multi-start L-BFGS-B optimization with
Nelder-Mead fallback.  Standard errors come from the observed Fisher
information (inverse of the negative Hessian at the MLE).

```{r fit-model}
fit_fn <- fit(model)
result <- fit_fn(df, n_starts = 5L)

# MLE estimates
coef(result)

# Standard errors
sqrt(diag(vcov(result)))

# 95% confidence intervals
confint(result)

# Log-likelihood, AIC, BIC
c(loglik = logLik(result), AIC = AIC(result), BIC = BIC(result))
```

```{r fit-summary}
summary(result)
```

## Comparison with true values

```{r compare-truth}
est <- coef(result)
se <- sqrt(diag(vcov(result)))
ci <- confint(result)

comparison <- data.frame(
  component = 1:3,
  true = theta_true,
  estimate = est,
  se = se,
  ci_lower = ci[, 1],
  ci_upper = ci[, 2],
  covered = theta_true >= ci[, 1] & theta_true <= ci[, 2]
)
comparison
```


# Permutation Symmetry

Under Scheme 0, all components are candidates for every observation.  The
log-likelihood is then **invariant under permutation** of the parameter
vector: swapping $\lambda_1 \leftrightarrow \lambda_2$ yields the same
log-likelihood value.  This means there are $m!$ equivalent global maxima.

```{r permutation-demo}
ll_fn <- loglik(model)

# Original parameter order
ll_original <- ll_fn(df, theta_true)

# Permuted parameters: swap components 1 and 3
theta_permuted <- theta_true[c(3, 1, 2)]
ll_permuted <- ll_fn(df, theta_permuted)

# Reversed order
theta_reversed <- rev(theta_true)
ll_reversed <- ll_fn(df, theta_reversed)

data.frame(
  ordering = c("original (0.5, 0.3, 0.2)",
               "permuted (0.2, 0.5, 0.3)",
               "reversed (0.2, 0.3, 0.5)"),
  loglik = c(ll_original, ll_permuted, ll_reversed),
  difference = c(0, ll_permuted - ll_original, ll_reversed - ll_original)
)
```

The log-likelihood values are identical (up to machine precision).  This
symmetry has two important consequences:

1. **Initialization matters.** The optimizer may converge to any of the $m!$
   equivalent modes.  The `default_init_exp()` function uses a **spread
   initialization** --- rates spaced as `lam0 * seq(0.5, 1.5, length.out = m)`
   --- to break the symmetry and consistently find the ascending-order mode.

2. **Sorted matching for Monte Carlo.** When evaluating estimator performance
   across replications, we sort both the estimates and the true parameters in
   ascending order before computing bias, RMSE, and coverage.  This resolves
   the label-switching problem.


# Monte Carlo Validation

We assess the MLE's finite-sample performance via Monte Carlo simulation.
For each replicate, we generate data from the true model, fit the MLE, and
record the estimates, standard errors, and confidence interval coverage.

```{r mc-precomputed, include=FALSE, eval=!run_long}
# When run_long = FALSE, we run a small demo instead of loading precomputed
# results.  The full simulation (R = 100, n = 300) takes several minutes.
```

```{r mc-full, eval=run_long, echo=run_long, cache=TRUE}
set.seed(2026)

R <- 100        # Monte Carlo replications
n_mc <- 300     # Sample size per replicate
alpha <- 0.05   # Significance level

theta_true_mc <- c(0.5, 0.3, 0.2)
theta_sorted <- sort(theta_true_mc)
m_mc <- length(theta_true_mc)

gen_mc <- rdata(model)
fit_mc <- fit(model)

# Storage
estimates <- matrix(NA, nrow = R, ncol = m_mc)
se_hat <- matrix(NA, nrow = R, ncol = m_mc)
ci_lower <- matrix(NA, nrow = R, ncol = m_mc)
ci_upper <- matrix(NA, nrow = R, ncol = m_mc)
converged <- logical(R)

for (r in seq_len(R)) {
  df_r <- gen_mc(theta = theta_true_mc, n = n_mc, p = 1)
  res_r <- tryCatch(fit_mc(df_r, n_starts = 10L), error = function(e) NULL)

  if (!is.null(res_r) && !any(is.na(coef(res_r)))) {
    # Sort estimates ascending for identifiability
    ord <- order(coef(res_r))
    estimates[r, ] <- coef(res_r)[ord]
    se_r <- sqrt(diag(vcov(res_r)))
    se_hat[r, ] <- se_r[ord]
    ci_r <- confint(res_r, level = 1 - alpha)
    ci_lower[r, ] <- ci_r[ord, 1]
    ci_upper[r, ] <- ci_r[ord, 2]
    converged[r] <- TRUE
  }
}

# Save for reproducibility
saveRDS(
  list(estimates = estimates, se_hat = se_hat,
       ci_lower = ci_lower, ci_upper = ci_upper,
       converged = converged, R = R, n_mc = n_mc,
       theta_sorted = theta_sorted, alpha = alpha),
  "precomputed_exp_parallel.rds"
)
```

```{r mc-small, eval=!run_long}
# Small demo: R = 5 replications for illustration (keeps build < 30 sec).
# Set run_long = TRUE above for the full R = 100 simulation.
set.seed(2026)

R <- 5
n_mc <- 50
alpha <- 0.05

theta_true_mc <- c(0.5, 0.3, 0.2)
theta_sorted <- sort(theta_true_mc)
m_mc <- length(theta_true_mc)

gen_mc <- rdata(model)
fit_mc <- fit(model)

estimates <- matrix(NA, nrow = R, ncol = m_mc)
se_hat <- matrix(NA, nrow = R, ncol = m_mc)
ci_lower <- matrix(NA, nrow = R, ncol = m_mc)
ci_upper <- matrix(NA, nrow = R, ncol = m_mc)
converged <- logical(R)

for (r in seq_len(R)) {
  df_r <- gen_mc(theta = theta_true_mc, n = n_mc, p = 1)
  res_r <- tryCatch(fit_mc(df_r, n_starts = 3L), error = function(e) NULL)

  if (!is.null(res_r) && !any(is.na(coef(res_r)))) {
    ord <- order(coef(res_r))
    estimates[r, ] <- coef(res_r)[ord]
    se_r <- sqrt(diag(vcov(res_r)))
    se_hat[r, ] <- se_r[ord]
    ci_r <- confint(res_r, level = 1 - alpha)
    ci_lower[r, ] <- ci_r[ord, 1]
    ci_upper[r, ] <- ci_r[ord, 2]
    converged[r] <- TRUE
  }
}
```

## Results

After sorting each replicate's estimates in ascending order (to match the
sorted true values $\lambda_{(1)} = 0.2 < \lambda_{(2)} = 0.3 <
\lambda_{(3)} = 0.5$), we compute bias, RMSE, and coverage:

```{r mc-results}
ok <- converged & complete.cases(estimates)
cat("Converged replications:", sum(ok), "of", R, "\n\n")

est_ok <- estimates[ok, , drop = FALSE]
se_ok <- se_hat[ok, , drop = FALSE]
ci_lo <- ci_lower[ok, , drop = FALSE]
ci_hi <- ci_upper[ok, , drop = FALSE]

bias <- colMeans(est_ok) - theta_sorted
rmse <- sqrt(colMeans((est_ok - matrix(theta_sorted, nrow = sum(ok),
                                        ncol = m_mc, byrow = TRUE))^2))
coverage <- colMeans(
  ci_lo <= matrix(theta_sorted, nrow = sum(ok), ncol = m_mc, byrow = TRUE) &
  ci_hi >= matrix(theta_sorted, nrow = sum(ok), ncol = m_mc, byrow = TRUE)
)
mean_se <- colMeans(se_ok)

mc_table <- data.frame(
  component = paste0("lambda_(", 1:m_mc, ")"),
  true = theta_sorted,
  mean_est = colMeans(est_ok),
  bias = bias,
  rmse = rmse,
  mean_se = mean_se,
  coverage_95 = coverage
)
mc_table
```

Key observations:

- **Bias** should be near zero for a correctly specified MLE.
- **RMSE** reflects finite-sample variability; the smallest rate
  ($\lambda_{(1)} = 0.2$) is typically estimated with larger relative error
  because it contributes least to the system density (its component is
  least likely to be the last to fail).
- **Coverage** near 0.95 validates the Hessian-based confidence intervals.
  Note that the SE ratio artifact --- Hessian-based SEs computed at the
  unsorted MLE, then reordered --- can cause mild under- or over-coverage.


# Observation Functors

The `observe_*` family of functions controls how system lifetimes are
recorded.  Each functor maps a true system lifetime to an observed record
(time, observation type, and optional interval bounds).

## Scheme 0: System lifetime with right-censoring

`observe_scheme0(tau)` is the default: systems failing before $\tau$ are
observed exactly, and those surviving past $\tau$ are right-censored.

```{r observe-scheme0}
# Different censoring times
obs_light <- observe_scheme0(tau = 20)   # Light censoring
obs_heavy <- observe_scheme0(tau = 5)    # Heavy censoring
obs_none  <- observe_scheme0()           # No censoring (tau = Inf)

# Example: system fails at t = 8
obs_light(8)   # exact
obs_heavy(8)   # right-censored at tau = 5
obs_none(8)    # exact
```

## Effect of censoring on estimation

Heavier right-censoring discards information about long-lived systems,
increasing estimator variance --- especially for the smallest rate (the
most reliable component, which is most likely to be the last survivor).

```{r censoring-effect}
set.seed(99)

tau_values <- c(5, Inf)

cat(sprintf("%-8s  %6s  %10s  %10s  %10s\n",
            "tau", "n_right", "est_lam1", "est_lam2", "est_lam3"))
cat(strrep("-", 56), "\n")

for (tau in tau_values) {
  obs_fn <- if (is.finite(tau)) observe_scheme0(tau) else NULL
  df_tau <- gen(theta = theta_true, n = 100,
                p = 1, observe = obs_fn)
  n_right <- sum(df_tau$omega == "right")

  res_tau <- tryCatch(
    fit_fn(df_tau, n_starts = 3L),
    error = function(e) NULL
  )

  if (!is.null(res_tau)) {
    est <- sort(coef(res_tau))
    cat(sprintf("%-8s  %6d  %10.4f  %10.4f  %10.4f\n",
                if (is.finite(tau)) as.character(tau) else "Inf",
                n_right, est[1], est[2], est[3]))
  }
}
```

## Scheme 1: Periodic inspection

Under `observe_scheme1(delta, tau)`, the system is inspected at times
$\delta, 2\delta, 3\delta, \ldots$  The failure time is known only to lie
within the inspection interval containing it (interval-censored).

```{r observe-scheme1}
obs_periodic <- observe_scheme1(delta = 2, tau = 30)

# System fails at t = 7.3 -> interval-censored to (6, 8]
obs_periodic(7.3)

# System survives past tau = 30 -> right-censored
obs_periodic(35)
```

```{r scheme1-demo}
set.seed(42)
df_s1 <- gen(theta = theta_true, n = 100,
             p = 1, observe = observe_scheme1(delta = 2, tau = 50))
table(df_s1$omega)

# Fit under interval censoring
res_s1 <- fit_fn(df_s1, n_starts = 3L)
sort(coef(res_s1))
```

Interval censoring retains more information than pure right-censoring at the
same study duration, because every failure is localized to a finite window
rather than lost entirely.


```{r cleanup, include = FALSE}
options(old_opts)
```
