---
title: "General k-out-of-n Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{General k-out-of-n Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>"
)
library(kofn)
set.seed(42)
old_opts <- options(digits = 4)
```

# The k Spectrum

A k-out-of-n system fails when k components have failed. The parameter k
controls how system-level observations censor individual component lifetimes:

| k | System type | Censoring at system failure |
|---|-------------|----------------------------|
| 1 | Series      | 1 exact, 0 left, 3 right   |
| 2 | 2-of-4      | 1 exact, 1 left, 2 right   |
| 3 | 3-of-4      | 1 exact, 2 left, 1 right   |
| 4 | Parallel    | 1 exact, 3 left, 0 right   |

As k increases from series to parallel, more component lifetimes are
left-censored (already failed) and fewer are right-censored (still alive).

The system density engine handles all cases via the critical-state
enumeration in `f_sys_general()`:

```{r density-all-k}
rates <- c(1, 0.8, 0.6, 0.4)
t_grid <- seq(0.01, 6, length.out = 200)

dens <- sapply(1:4, function(k) {
  sys <- kofn_system(k, 4)
  dists <- make_dists(rates, "exponential")
  f_sys_general(t_grid, sys, dists)
})
colnames(dens) <- paste0("k=", 1:4)

# Peak density time shifts right as k increases
peak_times <- t_grid[apply(dens, 2, which.max)]
data.frame(k = 1:4, peak_time = round(peak_times, 2))
```

Series systems (k=1) fail early; parallel systems (k=4) survive longer
because all components must fail.


# Estimation Across k

We generate data from 4-component exponential systems at each k value
and estimate rates using `fit_system()`.  We use a single dataset per k
(n = 300) to keep the vignette fast.

```{r estimation-across-k}
set.seed(42)
rates <- c(1, 0.8, 0.6, 0.4)
rates_sorted <- sort(rates)

k_table <- data.frame(
  k = 1:4,
  type = c("series", "2-of-4", "3-of-4", "parallel"),
  sum_est = NA_real_,
  sum_true = sum(rates),
  mean_rate_err = NA_real_,
  converged = NA
)

for (k in 1:4) {
  sys <- kofn_system(k, 4)
  df <- rdata_system(sys, par = rates, n = 300)
  res <- tryCatch(fit_system(df$t, sys, n_starts = 5L),
                  error = function(e) NULL)
  if (!is.null(res) && res$converged) {
    k_table$sum_est[k] <- round(sum(coef(res)), 2)
    k_table$mean_rate_err[k] <- round(mean(abs(sort(coef(res)) - rates_sorted)), 3)
    k_table$converged[k] <- TRUE
  } else {
    k_table$converged[k] <- FALSE
  }
}
k_table
```

Key observations:

- **Series (k=1)** typically gives the best individual rate estimation.
  For exponential series, the system hazard equals $\sum_j \lambda_j$,
  so the sum of rates is directly identifiable from the system failure rate.
- **Parallel (k=4)** is the hardest case. The system density is a
  symmetric function of the rates (invariant under permutation), creating
  $m!$ equivalent modes on the likelihood surface.  The optimizer may
  converge to a spurious symmetric solution where all rates are equal.
- **Intermediate k** values balance left and right censoring, but
  permutation symmetry still limits identifiability of individual rates
  from system-level data alone.


# Weibull k-out-of-n

The Weibull model works for any k, not just the parallel special case.
Here we generate from a 2-out-of-3 system and recover the parameters.

```{r weibull-kofn}
set.seed(123)
wei_model <- kofn(k = 2, m = 3, family = "weibull")
print(wei_model)

# True parameters: (shape_1, scale_1, shape_2, scale_2, shape_3, scale_3)
theta_wei <- c(1.5, 2.0,  2.0, 3.0,  1.0, 1.5)
gen_wei <- rdata(wei_model)
df_wei <- gen_wei(theta_wei, n = 150)

# Fit
fit_wei <- fit(wei_model)
res_wei <- fit_wei(df_wei, n_starts = 5L)
```

```{r weibull-results}
est <- coef(res_wei)
true_shapes <- theta_wei[seq(1, 6, by = 2)]
true_scales <- theta_wei[seq(2, 6, by = 2)]
est_shapes  <- est[seq(1, 6, by = 2)]
est_scales  <- est[seq(2, 6, by = 2)]

# Sort by shape for identifiability comparison
ord_true <- order(true_shapes)
ord_est  <- order(est_shapes)

data.frame(
  component   = 1:3,
  true_shape  = true_shapes[ord_true],
  est_shape   = est_shapes[ord_est],
  true_scale  = true_scales[ord_true],
  est_scale   = est_scales[ord_est]
)
```

The direct MLE recovers both shape and scale parameters via the general
system density engine, which enumerates critical states for each
component at each time point.


# Bridge and Custom Systems

The `kofn()` constructor also accepts arbitrary coherent systems via
the `system` argument. The standard bridge system has 5 components
with non-k-out-of-n structure.

```{r bridge-system}
bridge <- bridge_system()
print(bridge)

# System signature shows which order statistic determines system failure
system_signature(bridge)
```

```{r bridge-estimation}
set.seed(77)
bridge_model <- kofn(system = bridge, family = "exponential")
print(bridge_model)

rates_bridge <- c(1.0, 0.8, 0.6, 1.2, 0.9)
df_bridge <- rdata_system(bridge, par = rates_bridge, n = 200)

res_bridge <- fit_system(df_bridge$t, bridge, n_starts = 5L)
est_bridge <- coef(res_bridge)

data.frame(
  component = 1:5,
  true_rate = rates_bridge,
  estimate  = est_bridge,
  se        = sqrt(diag(vcov(res_bridge)))
)
```

The `fit_system()` function works for any `coherent_system` object --
it uses the general density engine regardless of system structure.


# Identifiability

The amount of information available about individual component parameters
depends on the system structure:

- **Series (k=1)**: Each observation is right-censored for m-1 components
  and exact for one. Without masking information (which component failed),
  only the total hazard $\sum_j \lambda_j$ is identifiable for
  exponential components. The `maskedcauses` package handles this case
  with candidate-set data.

- **Parallel (k=n)**: All components have failed by system failure time,
  but we observe only the maximum. Permutation symmetry means the
  likelihood is invariant to relabeling -- there are $m!$ equivalent
  modes. The optimizer's spread initialization and ascending-order
  convention resolve this for estimation.

- **Intermediate k**: A mix of left- and right-censored components.
  More balanced censoring generally improves identifiability of
  individual rates, though permutation symmetry persists.

```{r identifiability-demo}
# Series: log-likelihood is identical under permutation of rates
sys_series <- kofn_system(1, 3)
set.seed(1)
df_s <- rdata_system(sys_series, par = c(1, 2, 3), n = 100)

ll_1 <- loglik_system(df_s$t, sys_series, c(1, 2, 3))
ll_2 <- loglik_system(df_s$t, sys_series, c(3, 1, 2))
c(original = ll_1, permuted = ll_2, diff = ll_1 - ll_2)
```

For series systems with identical-distribution components, only the sum of
rates matters. For general k-out-of-n, the system structure creates
asymmetric censoring patterns that can break some symmetries, but
identifiability of individual rates remains fundamentally limited under
Scheme 0.

```{r cleanup, include = FALSE}
options(old_opts)
```
