---
title: "Getting started with kofn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with kofn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(42)
```

```{r setup}
library(kofn)
```

## What `kofn` does

`kofn` is a maximum likelihood estimation toolkit for **k-out-of-n
systems**: reliability systems that fail when `k` of `m` components
have failed. `k = 1` is a series system (first failure ends it), `k =
m` is a parallel system (all must fail), and intermediate values
interpolate between them.

The package focuses exclusively on inference: log-likelihood, score,
Hessian, fit, observation schemes (right/left/interval/periodic),
masked cause-of-failure, and Fisher information comparison. The
underlying data-generating process (DGP) and topology queries are
delegated to `dist.structure`, which kofn imports.

## Your first fit

```{r}
# A 2-of-3 exponential system: system fails when 2 of 3 components fail.
model <- kofn(k = 2, m = 3, family = "exponential")

# Generate 200 complete (Scheme 0) observations.
gen <- rdata(model)
df <- gen(theta = c(1, 2, 3), n = 200)

# Fit via maximum likelihood.
fitter <- fit(model)
result <- fitter(df, n_starts = 3)
result$converged
coef(result)
```

For exponential parallel systems (k = m), only the sum of rates is
identifiable from system-level data alone; individual rates are not.
The sum typically recovers well:

```{r}
sum(coef(result))
sum(c(1, 2, 3))
```

## Convention

kofn uses the **:F convention**: `k` counts component failures that
trigger system failure.

- `kofn(k = 1, m)` = series (first failure ends the system).
- `kofn(k = m, m)` = parallel (system survives until the m-th failure).
- `kofn(k, m)` for intermediate k = k-out-of-m systems.

(dist.structure internally uses the :G convention, which is the dual.
kofn handles the conversion for you; you never need to think about
:G.)

## Observation schemes

Real experiments rarely observe system failure exactly. kofn provides
composable observation functors for censoring and periodic inspection:

```{r}
# Right-censored: systems that fail after tau = 2 get recorded at t = 2
df_cens <- gen(c(1, 2, 3), n = 200,
               observe = observe_right_censor(tau = 2))
table(df_cens$omega)
```

The log-likelihood handles mixed observation types automatically:

```{r}
ll <- loglik(model)
ll(df_cens, c(1, 2, 3))  # finite; right-censored obs contribute log S(t)
```

Other functors: `observe_left_censor`, `observe_interval_censor`,
`observe_periodic`, `observe_mixture`. See the observation-schemes
vignette.

## Weibull components

Weibull component lifetimes work analogously, with parameters
interleaved as `(shape_1, scale_1, shape_2, scale_2, ...)`:

```{r}
model_wei <- kofn(k = 2, m = 2, family = "weibull", method = "em")
gen_wei <- rdata(model_wei)
df_wei <- gen_wei(theta = c(1.5, 2.0, 2.0, 3.0), n = 300)
result_wei <- fit(model_wei)(df_wei, n_starts = 3)
result_wei$shapes
result_wei$scales
```

For parallel Weibull systems (`k = m`), `method = "em"` uses an EM
algorithm treating the identity of the last-failing component as
latent. For general k, use `method = "mle"` (direct L-BFGS-B).

## Periodic inspection (Scheme 1)

When only periodic snapshots are available, use
`loglik_scheme1`/`fit_scheme1`:

```{r}
s1gen <- rdata_scheme1(model)
df_s1 <- s1gen(theta = c(1, 2, 3), n = 200, delta = 0.5)
ll_s1 <- loglik_scheme1(model)
ll_s1(df_s1, c(1, 2, 3))
```

## Masked cause-of-failure

When you observe system failure and a candidate set (superset of the
truly-failed components) rather than knowing which components failed:

```{r}
rgen <- rdata_masked(model)
df_masked <- rgen(theta = c(1, 2, 3), n = 100, p_mask = 0.3)
ll_masked <- loglik_masked(model)
ll_masked(df_masked, c(1, 2, 3))
```

## Fisher information comparison

How much precision do you gain from each observation scheme? kofn
provides a built-in comparison:

```{r, eval = FALSE}
res <- compare_fisher_info(
  rates = c(0.5, 0.3), n = 50, delta = 1.0, n_rep = 10,
  family = "exponential"
)
res$median_det
```

This runs `n_rep` replications of each scheme and reports the median
determinant of the observed information matrix per scheme.

## What's next

See the other vignettes for deeper coverage:

- **[Exponential parallel](exponential-parallel.html)**: the IE
  expansion fast path for parallel exponentials.
- **[Observation schemes](observation-schemes.html)**: composable
  observation functors for complex monitoring setups.
- **[Weibull EM](weibull-em.html)**: the EM algorithm details.
- **[Periodic inspection](periodic-inspection.html)**: Scheme 1
  workflows.
- **[dist.structure integration](dist-structure-integration.html)**:
  how kofn delegates DGP and topology to dist.structure; migration
  notes for v0.2.0 users.
- **[Ecosystem](ecosystem.html)**: how kofn fits with
  `likelihood.model`, `algebraic.mle`, and `dist.structure`.
