---
title: "Coherent Systems and the k-out-of-n Spectrum"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coherent Systems and the k-out-of-n Spectrum}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(42)
```

The `kofn` package provides infrastructure for component lifetime estimation
from system-level data, framed as a censoring problem. At its foundation is a
**coherent system engine** that represents arbitrary system structures via
minimal path sets, computes dual cut-set representations, and determines the
censoring pattern that each structure imposes on component lifetimes.

This vignette walks through the system engine and shows how system structure
affects estimation quality. For Weibull-specific EM estimation, see the
`weibull-em` vignette.

```{r load-package}
library(kofn)
```


## 1. Coherent systems

A **coherent system** is defined by its minimal path sets: the smallest sets of
components whose simultaneous functioning guarantees system functioning. The
`coherent_system()` constructor takes a list of integer vectors (each a minimal
path set) and precomputes the dual minimal cut sets eagerly.

```{r coherent-system-basic}
# A custom system: works if {1,2} or {2,3} are both functioning
sys <- coherent_system(list(c(1, 2), c(2, 3)), m = 3)
sys
```

The package provides constructors for all standard system types:

```{r standard-constructors}
# Series: single path containing all components (all must work)
series_system(4)

# Parallel: one path per component (any one suffices)
parallel_system(4)

# k-out-of-n: works when at least k of n components function
kofn_system(2, 4)

# Bridge: the classic reliability textbook system (5 components)
bridge_system()
```

A **k-out-of-n system** unifies series and parallel as endpoints of a spectrum.
Setting k = 1 gives series (one failure kills the system), while k = n gives
parallel (all must fail). Every intermediate k value defines a distinct
censoring structure on the component lifetimes.


## 2. Path sets and cut sets

Path sets and cut sets are dual representations of the same system. A **minimal
cut set** is a smallest set of components whose simultaneous failure causes
system failure. The `min_cuts()` accessor retrieves them:

```{r path-cut-duality}
sys_24 <- kofn_system(2, 4)

# Minimal path sets: all pairs (since any 2 functioning = system works)
cat("Path sets:\n")
for (p in sys_24$min_paths) cat("  {", paste(p, collapse = ", "), "}\n")

# Minimal cut sets: all triples (need 3 failures to kill the system)
cat("\nCut sets:\n")
for (cc in min_cuts(sys_24)) cat("  {", paste(cc, collapse = ", "), "}\n")
```

The cut sets are computed via the **Berge transversal algorithm**: a minimal
transversal (hitting set) of the path set hypergraph is exactly a minimal cut
set. The algorithm iteratively extends partial transversals by elements of each
path set, then prunes non-minimal results:

```{r berge-demo}
# Bridge system: 4 paths, complex structure
br <- bridge_system()
cat("Bridge paths:\n")
for (p in br$min_paths) cat("  {", paste(p, collapse = ", "), "}\n")

cat("\nBridge cuts (computed via Berge transversal):\n")
for (cc in min_cuts(br)) cat("  {", paste(cc, collapse = ", "), "}\n")
```

The path-cut duality extends to lifetimes. Given path sets $P_1, \ldots, P_r$
and cut sets $C_1, \ldots, C_s$:

$$T_{\mathrm{sys}} = \max_P \min_{j \in P} T_j = \min_C \max_{j \in C} T_j$$

The min-of-max (cut-based) representation is used internally by
`system_lifetime()` because it is computationally more natural for determining
when the system fails.


## 3. Structure function and lifetime

The **structure function** $\phi(\mathbf{x})$ maps a logical component state
vector to a system state. The system works if and only if at least one minimal
path set is fully operational:

```{r structure-function}
sys <- kofn_system(2, 3)

# 2-of-3: need at least 2 functioning components
phi(sys, c(TRUE, TRUE, FALSE))   # two up -> works
phi(sys, c(TRUE, FALSE, FALSE))  # only one up -> fails
phi(sys, c(TRUE, TRUE, TRUE))    # all up -> works
```

The `system_lifetime()` function computes $T_{\mathrm{sys}}$ from component
lifetimes using the min-of-max cut-set formula:

```{r system-lifetime}
times <- c(1.5, 2.3, 0.8)

# Parallel (k=3): T = max(T_j) = 2.3
system_lifetime(parallel_system(3), times)

# Series (k=1): T = min(T_j) = 0.8
system_lifetime(series_system(3), times)

# 2-of-3: T = T_(2) = 1.5 (second order statistic)
system_lifetime(kofn_system(2, 3), times)
```

For a k-out-of-n system with i.i.d. components, the system lifetime is the
k-th order statistic $T_{(k)}$. For non-i.i.d. components or general coherent
systems, `system_lifetime()` handles the general case via the cut-set
computation.


## 4. System signatures

The **system signature** is the probability vector $\mathbf{s}$ where
$s_k = P(T_{\mathrm{sys}} = T_{(k)})$ under i.i.d. continuous component
lifetimes. It depends only on the system structure, not on the component
distribution.

```{r signatures}
# Series: always fails at the first order statistic
system_signature(series_system(4))

# Parallel: always fails at the last order statistic
system_signature(parallel_system(4))

# 2-out-of-4: spread across middle order statistics
system_signature(kofn_system(2, 4))

# Bridge: non-trivial mixture
system_signature(bridge_system())
```

The signature tells us which order statistics contribute to the system lifetime.
This directly affects estimation quality: signatures concentrated on extreme
order statistics (series or parallel) carry less information about individual
component parameters than signatures spread across the middle.

```{r signature-comparison}
# Compare signatures for k-out-of-5 systems
sig_table <- data.frame(
  order_stat = paste0("T_(", 1:5, ")"),
  k1_series  = system_signature(kofn_system(1, 5)),
  k2         = system_signature(kofn_system(2, 5)),
  k3         = system_signature(kofn_system(3, 5)),
  k4         = system_signature(kofn_system(4, 5)),
  k5_parallel = system_signature(kofn_system(5, 5))
)
knitr::kable(sig_table, digits = 3,
             caption = "System signatures for k-out-of-5 systems")
```


## 5. Censoring patterns

The key insight of the censoring framework: observing a system lifetime creates
structured censoring on the individual component lifetimes. The
`system_censoring()` function determines, for each component, whether its
lifetime is observed exactly, left-censored (failed before the system), or
right-censored (still alive at system failure):

```{r censoring-basic}
# Parallel system (k=m): system fails when ALL components fail
sys_par <- parallel_system(3)
times <- c(1.5, 2.3, 0.8)
cens <- system_censoring(sys_par, times)
cat("System lifetime:", cens$T_sys, "\n")
cat("Critical component:", cens$critical, "\n")
cat("Censoring status:", paste(cens$status, collapse = ", "), "\n")
```

For a k-out-of-n system with lifetime $T = T_{(k)}$:

- **1 component** observed exactly (the critical component, $T_j = T$)
- **k - 1 components** left-censored at $T$ (failed before the system)
- **n - k components** right-censored at $T$ (still alive at system failure)

Let us verify this pattern across the k-out-of-5 spectrum:

```{r censoring-spectrum}
m <- 5
# Use well-separated times so the pattern is unambiguous
times <- c(1.0, 2.0, 3.0, 4.0, 5.0)

cens_table <- data.frame(
  k = 1:5,
  system_type = c("series", "2-of-5", "3-of-5", "4-of-5", "parallel"),
  T_sys = NA_real_,
  exact = NA_integer_,
  left = NA_integer_,
  right = NA_integer_
)

for (i in 1:5) {
  sys <- kofn_system(i, m)
  cens <- system_censoring(sys, times)
  cens_table$T_sys[i] <- cens$T_sys
  cens_table$exact[i] <- sum(cens$status == "exact")
  cens_table$left[i]  <- sum(cens$status == "left")
  cens_table$right[i] <- sum(cens$status == "right")
}

knitr::kable(cens_table,
             caption = "Censoring pattern for k-out-of-5 systems")
```

The table confirms the general rule: k = 1 (series) gives 1 exact + 0 left + 4 right,
k = 5 (parallel) gives 1 exact + 4 left + 0 right, and intermediate values interpolate
between these extremes. This is the **censoring equivalence** that unifies
parallel and series systems as endpoints of a spectrum.


## 6. Estimation across system structures

How does the system structure affect our ability to estimate component
parameters? We generate data from k-out-of-5 exponential systems for each k
value and compare estimation quality.

```{r estimation-comparison}
m <- 5
true_rates <- c(0.5, 0.8, 1.0, 1.3, 1.8)
n <- 200
results <- data.frame(k = 1:5, system_type = c("series", "2-of-5", "3-of-5", "4-of-5", "parallel"),
                      rmse = NA_real_, stringsAsFactors = FALSE)

for (k in 1:5) {
  sys <- kofn_system(k, m)
  dat <- rdata_system(sys, par = true_rates, family = "exponential", n = n)
  res <- fit_system(dat$t, sys, family = "exponential", n_starts = 3)
  if (res$convergence == 0) {
    results$rmse[k] <- sqrt(mean((sort(res$par) - sort(true_rates))^2))
  }
}

knitr::kable(results, digits = 4,
             caption = "Single-sample estimation quality by system structure")
```

A single sample gives only a rough indication. A small Monte Carlo study gives
more reliable RMSE estimates:

```{r monte-carlo-comparison, fig.height = 4.5}
m <- 3
true_rates <- c(0.5, 1.0, 2.0)
n <- 150
n_rep <- 30

mc_results <- data.frame(
  k = integer(), replicate = integer(),
  rmse = numeric(), stringsAsFactors = FALSE
)

for (k in 1:3) {
  sys <- kofn_system(k, m)
  for (rep in seq_len(n_rep)) {
    dat <- rdata_system(sys, par = true_rates, family = "exponential", n = n)
    res <- fit_system(dat$t, sys, family = "exponential", n_starts = 3)
    if (res$convergence == 0) {
      est_sorted <- sort(res$par)
      rmse <- sqrt(mean((est_sorted - sort(true_rates))^2))
      mc_results <- rbind(mc_results, data.frame(
        k = k, replicate = rep, rmse = rmse, stringsAsFactors = FALSE
      ))
    }
  }
}

boxplot(rmse ~ k, data = mc_results,
        main = "Estimation RMSE by System Structure (k-out-of-3)",
        xlab = "k (1 = series, 3 = parallel)",
        ylab = "RMSE of sorted rate estimates",
        col = "lightblue", border = "steelblue")
```

The censoring pattern explains the differences. Under k = 1 (series), most
component lifetimes are right-censored: we know they survived past the system
failure but not how long. Under k = 3 (parallel for m=3), most are left-censored:
we know they failed before the system but not exactly when. Intermediate values
of k balance left and right censoring, potentially yielding the most
information about individual component lifetimes.


## 7. Beyond k-out-of-n

The coherent system engine is not limited to k-out-of-n systems. Any system
that can be described by its minimal path sets is supported. Two additional
constructors demonstrate this generality.

### Bridge systems

The **bridge system** is a classic 5-component reliability structure with four
minimal path sets. It cannot be represented as any k-out-of-n system:

```{r bridge-signature}
br <- bridge_system()
cat("Bridge signature:\n")
print(round(system_signature(br), 4))
```

The signature has support on $T_{(2)}$, $T_{(3)}$, and $T_{(4)}$ -- a pattern
impossible for any k-out-of-n system, which always concentrates on a single
order statistic plus its neighbors.

We can generate data and estimate parameters for the bridge system using the
same engine:

```{r bridge-estimation}
br <- bridge_system()
true_rates <- c(0.5, 0.8, 1.0, 1.3, 1.6)
n <- 200

dat <- rdata_system(br, par = true_rates, family = "exponential", n = n)
res <- fit_system(dat$t, br, family = "exponential", n_starts = 5)

if (res$convergence == 0) {
  se_vals <- if (length(res$se) == 5 && !all(is.na(res$se))) {
    sort(res$se)
  } else {
    rep(NA_real_, 5)
  }
  comp_df <- data.frame(
    component = 1:5,
    true_rate = sort(true_rates),
    estimated = sort(res$par),
    se = se_vals
  )
  knitr::kable(comp_df, digits = 3,
               caption = "Bridge system estimation (n = 200, sorted)")
}
```

### Consecutive-k systems

A **consecutive-k-out-of-n system** fails when k consecutive components all
fail. This models physical systems where failures propagate along a chain
(pipelines, relay networks, communication links):

```{r consecutive-k}
# Consecutive-2-out-of-5: fails if any two adjacent components fail
cons <- consecutive_k_system(2, 5)
cons

# Consecutive system signature
cat("Consecutive-2-of-5 signature:\n")
print(round(system_signature(cons), 4))
```

The censoring pattern of a consecutive-k system depends on the spatial
arrangement of failures, not just their count. This makes it structurally
different from ordinary k-out-of-n systems:

```{r consecutive-censoring}
cons <- consecutive_k_system(2, 5)

# Scenario 1: non-adjacent failures (system survives longer)
times1 <- c(1.0, 5.0, 2.0, 5.0, 3.0)  # failures at positions 1,3,5 first
cens1 <- system_censoring(cons, times1)
cat("Non-adjacent early failures:\n")
cat("  T_sys =", cens1$T_sys, "\n")
cat("  Status:", paste(cens1$status, collapse = ", "), "\n\n")

# Scenario 2: adjacent failures (system fails earlier)
times2 <- c(5.0, 1.0, 2.0, 5.0, 5.0)  # components 2,3 fail early
cens2 <- system_censoring(cons, times2)
cat("Adjacent early failures:\n")
cat("  T_sys =", cens2$T_sys, "\n")
cat("  Status:", paste(cens2$status, collapse = ", "), "\n")
```

### General coherent systems

Any coherent system can be specified directly via `coherent_system()`. Here is a
system with an asymmetric structure:

```{r custom-system}
# Custom: works if {1,2,3} all work, OR if {3,4} both work
custom <- coherent_system(list(c(1, 2, 3), c(3, 4)), m = 4)
custom

cat("Signature:\n")
print(round(system_signature(custom), 4))
```

The system density engine (`f_sys_general`, `S_sys_general`) and the fitting
engine (`fit_system`) handle all coherent systems uniformly. The computational
cost scales as $O(2^m)$ per time point, which is practical for systems with up
to about 15 components.

```{r general-fit}
custom <- coherent_system(list(c(1, 2, 3), c(3, 4)), m = 4)
true_rates <- c(0.5, 1.0, 1.5, 2.0)

dat <- rdata_system(custom, par = true_rates, family = "exponential", n = 200)
res <- fit_system(dat$t, custom, family = "exponential", n_starts = 5)

if (res$convergence == 0) {
  cat("True rates (sorted): ", paste(sort(true_rates), collapse = ", "), "\n")
  cat("Estimated (sorted):  ", paste(round(sort(res$par), 3), collapse = ", "), "\n")
}
```

## Summary

The coherent system engine in `kofn` provides a unified framework for:

1. **Representing** arbitrary system structures via minimal path sets, with
   automatic computation of dual cut sets via the Berge transversal algorithm.

2. **Computing** system lifetimes, structure functions, and system signatures
   for any coherent system.

3. **Determining** the censoring pattern that each system structure imposes on
   component lifetimes -- the foundation of the censoring-based estimation
   framework.

4. **Estimating** component lifetime parameters from system-level data,
   regardless of system structure, using the general system density engine.

The k-out-of-n family provides a natural one-parameter spectrum from series
(k = 1) to parallel (k = n), with each value of k defining a distinct balance of
left and right censoring on the component lifetimes. General coherent systems
(bridge, consecutive-k, custom) are handled by the same machinery, enabling
estimation under any system architecture.
