---
title: "Observation Schemes: Resolving Information Asymmetry via Periodic Inspection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Observation Schemes: Resolving Information Asymmetry via Periodic Inspection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
library(kofn)

# Heavy simulations are off by default so the vignette builds in <30 seconds.
# Set run_long <- TRUE to reproduce all Monte Carlo results.
run_long <- FALSE
```

## Introduction

When we observe only the lifetime of a parallel system --- the time until
the *last* component fails --- the system lifetime
$T = \max(T_1, \ldots, T_m)$ is a single number, but we want to recover
$m$ individual component lifetime distributions. Components that fail
quickly ("fast" components) have $F_j(T) \approx 1$ at the system
failure time, contributing almost no curvature to the likelihood.
Slow components dominate the system density and are estimated well.
This **information asymmetry** is the central practical obstacle.

This vignette shows how **periodic inspection** --- checking which
components have failed at regular intervals --- resolves this asymmetry
at surprisingly low cost.


## 1. The observation scheme hierarchy

The `kofn` package defines three observation schemes, ordered from least
to most informative (**Scheme 2 > Scheme 1 > Scheme 0**):

| Scheme | What is observed | Constructor |
|--------|------------------|-------------|
| Scheme 0 | System lifetime only (black box) | `observe_right_censor()` |
| Scheme 1 | System lifetime + periodic component inspections | `observe_periodic(delta)` |
| Scheme 2 | All component lifetimes exactly (complete data) | `observe_exact()` |

Scheme 1 occupies the practical middle ground. At each inspection time
$\ell\delta$, we record which components have already failed. This
requires only periodic access to the system.

```{r observation-functors}
# Scheme 0: system lifetime only (optionally right-censored at tau)
obs0 <- observe_right_censor(tau = 100)
obs0(42.7)    # exact observation
obs0(150.0)   # right-censored at tau = 100

# Scheme 1: periodic inspection with interval width delta = 5
obs1 <- observe_periodic(delta = 5, tau = 100)
obs1(42.7)    # failure in interval [40, 45)
obs1(150.0)   # right-censored

# Scheme 2: complete data (trivial)
obs2 <- observe_exact()
obs2(42.7)    # exact observation, always
```


## 2. The information asymmetry problem

To see why Scheme 0 struggles, consider a two-component Weibull parallel
system with parameters:

- Component 1 (fast): shape $\alpha_1 = 1.5$, scale $\beta_1 = 2.0$
- Component 2 (slow): shape $\alpha_2 = 2.0$, scale $\beta_2 = 3.0$

Component 1 has a median lifetime of about 1.4 time units; Component 2's
median is about 2.5. At the system failure time $T = \max(T_1, T_2)$,
the fast component has almost surely already failed --- its CDF
$F_1(T) \approx 1$, so the likelihood surface is nearly flat with respect
to Component 1's parameters.

```{r asymmetry-demo}
set.seed(42)

# True parameters: (shape_1, scale_1, shape_2, scale_2)
theta_true <- c(1.5, 2.0, 2.0, 3.0)

# Create the Weibull parallel system model (EM for Scheme 0)
model_em <- kofn(k = 2, m = 2, family = "weibull", method = "em")

# Generate Scheme 0 data (system lifetimes only)
gen <- rdata(model_em)
df0 <- gen(theta_true, n = 300)

cat("System lifetime summary:\n")
summary(df0$t)

# At the median system lifetime, how much has each component's CDF accumulated?
t_med <- median(df0$t)
cat(sprintf("\nAt median system time %.2f:\n", t_med))
cat(sprintf("  F_1(t) = %.4f  (fast component: nearly saturated)\n",
            pweibull(t_med, shape = 1.5, scale = 2.0)))
cat(sprintf("  F_2(t) = %.4f  (slow component: still informative)\n",
            pweibull(t_med, shape = 2.0, scale = 3.0)))
```

With $F_1(T) \approx 1$, Component 1's contribution to the system
density is a tiny fraction of the total, and the likelihood carries
almost no information about $\alpha_1$ and $\beta_1$. Fitting under
Scheme 0 confirms this:

```{r scheme0-fit}
# Fit under Scheme 0 using the EM algorithm
solver <- fit(model_em)
mle0 <- solver(df0, n_starts = 3L)

cat("Scheme 0 estimates vs truth:\n")
cat(sprintf("  Component 1 (fast): shape = %.3f (true: 1.500), scale = %.3f (true: 2.000)\n",
            mle0$par[1], mle0$par[2]))
cat(sprintf("  Component 2 (slow): shape = %.3f (true: 2.000), scale = %.3f (true: 3.000)\n",
            mle0$par[3], mle0$par[4]))

# Standard errors from the variance-covariance matrix
se_vec <- tryCatch({
  V <- vcov(mle0)
  if (is.null(V)) rep(NA_real_, length(coef(mle0))) else sqrt(diag(V))
}, error = function(e) rep(NA_real_, length(coef(mle0))))
if (!any(is.na(se_vec))) {
  cat(sprintf("\n  SE(shape_1) = %.3f,  SE(shape_2) = %.3f\n",
              se_vec[1], se_vec[3]))
  cat(sprintf("  SE ratio (fast/slow): %.1fx\n", se_vec[1] / se_vec[3]))
}
```

The SE ratio reflects a genuine information gap, not an algorithm deficiency.


## 3. Scheme 1: Periodic inspection

Under Scheme 1, we supplement system-level observation with periodic
inspections. At each inspection time $\ell\delta$ (for $\ell = 1, 2, \ldots$),
we check each component and record whether it has failed. This localizes
each component's failure time to an inspection interval
$[a_j^{-}, a_j^{+})$ where $a_j^{+} - a_j^{-} = \delta$.

### The likelihood

The likelihood for a single observation under Scheme 1 combines two
pieces of information:

$$
L_i(\theta) = \underbrace{f_{\text{sys}}(t_i; \theta)}_{\text{exact system time}} \times \prod_{j=1}^{m} \underbrace{\left[F_j(a_{ij}^{+}) - F_j(a_{ij}^{-})\right]}_{\text{interval-censored component } j}
$$

where $f_{\text{sys}}(t) = \sum_{j=1}^{m} f_j(t) \prod_{k \neq j} F_k(t)$
is the parallel system density and $[a_{ij}^{-}, a_{ij}^{+})$ is the
inspection interval containing component $j$'s failure time.

Even a wide interval provides information: knowing that the fast
component failed in $[0, \delta)$ versus $[\delta, 2\delta)$ constrains
the hazard function in a way that the system-level observation alone
cannot. The `kofn` package provides dedicated functions for Scheme 1:

```{r scheme1-demo}
# Generate data under Scheme 1 with delta = 0.5
model_wei <- kofn(k = 2, m = 2, family = "weibull")
gen_s1 <- rdata_scheme1(model_wei)
df1 <- gen_s1(theta_true, n = 300, delta = 0.5)

cat("Scheme 1 data (first 6 observations):\n")
print(head(df1))

cat("\nColumn meanings:\n")
cat("  t:            exact system failure time\n")
cat("  comp_lower_j: lower bound of component j's inspection interval\n")
cat("  comp_upper_j: upper bound of component j's inspection interval\n")
```

```{r scheme1-fit}
# Fit under Scheme 1
solver_s1 <- fit_scheme1(model_wei)
mle1 <- solver_s1(df1, n_starts = 3L)

cat("Scheme 1 (delta = 0.5) estimates vs truth:\n")
cat(sprintf("  Component 1 (fast): shape = %.3f (true: 1.500), scale = %.3f (true: 2.000)\n",
            mle1$par[1], mle1$par[2]))
cat(sprintf("  Component 2 (slow): shape = %.3f (true: 2.000), scale = %.3f (true: 3.000)\n",
            mle1$par[3], mle1$par[4]))
```


## 4. Scheme 0 vs Scheme 1 comparison

The real test is a controlled comparison using the same true parameters.
We generate data from a common set of component lifetimes and fit under
both schemes.

```{r comparison-single, fig.height = 6}
set.seed(123)

n <- 300
m <- 2
alpha_true <- c(1.5, 2.0)
beta_true  <- c(2.0, 3.0)
theta_true <- c(alpha_true[1], beta_true[1], alpha_true[2], beta_true[2])
delta <- 0.5

# Generate component lifetimes directly
comp_times <- matrix(0, nrow = n, ncol = m)
for (j in seq_len(m)) {
  comp_times[, j] <- rweibull(n, shape = alpha_true[j], scale = beta_true[j])
}
sys_times <- apply(comp_times, 1, max)

# --- Scheme 0: system lifetime only ---
df_s0 <- data.frame(t = sys_times)
model_em <- kofn(k = 2, m = 2, family = "weibull", method = "em")
solver_em <- fit(model_em)
mle_s0 <- solver_em(df_s0, n_starts = 3L)

# --- Scheme 1: add inspection intervals ---
df_s1 <- data.frame(t = sys_times)
for (j in seq_len(m)) {
  df_s1[[paste0("comp_lower_", j)]] <- floor(comp_times[, j] / delta) * delta
  df_s1[[paste0("comp_upper_", j)]] <- floor(comp_times[, j] / delta) * delta + delta
}
model_mle <- kofn(k = 2, m = 2, family = "weibull", method = "mle")
solver_s1 <- fit_scheme1(model_mle)
mle_s1 <- solver_s1(df_s1, n_starts = 3L)

# --- Display results ---
cat("=== Single-sample comparison (n = 300) ===\n\n")
results <- data.frame(
  Parameter = c("shape_1", "scale_1", "shape_2", "scale_2"),
  Truth     = theta_true,
  Scheme_0  = round(mle_s0$par, 3),
  Scheme_1  = round(mle_s1$par, 3)
)
results$Error_S0 <- round(abs(results$Scheme_0 - results$Truth), 3)
results$Error_S1 <- round(abs(results$Scheme_1 - results$Truth), 3)
print(results, row.names = FALSE)
```

### Monte Carlo comparison

```{r mc-comparison, eval = run_long}
# Monte Carlo comparison: run_long = TRUE to execute
set.seed(2024)
n_rep <- 200
n <- 300
delta <- 0.5

est_s0 <- matrix(NA, nrow = n_rep, ncol = 4)
est_s1 <- matrix(NA, nrow = n_rep, ncol = 4)
conv_s0 <- logical(n_rep)
conv_s1 <- logical(n_rep)

for (r in seq_len(n_rep)) {
  # Generate common component lifetimes
  comp_times <- matrix(0, nrow = n, ncol = m)
  for (j in seq_len(m)) {
    comp_times[, j] <- rweibull(n, shape = alpha_true[j], scale = beta_true[j])
  }
  sys_times <- apply(comp_times, 1, max)

  # Scheme 0
  df_s0 <- data.frame(t = sys_times)
  res0 <- tryCatch(solver_em(df_s0, n_starts = 3L),
                    error = function(e) NULL)
  if (!is.null(res0) && res0$converged) {
    est_s0[r, ] <- res0$par
    conv_s0[r] <- TRUE
  }

  # Scheme 1
  df_s1 <- data.frame(t = sys_times)
  for (j in seq_len(m)) {
    df_s1[[paste0("comp_lower_", j)]] <- floor(comp_times[, j] / delta) * delta
    df_s1[[paste0("comp_upper_", j)]] <- floor(comp_times[, j] / delta) * delta + delta
  }
  res1 <- tryCatch(solver_s1(df_s1, n_starts = 3L),
                    error = function(e) NULL)
  if (!is.null(res1) && res1$converged) {
    est_s1[r, ] <- res1$par
    conv_s1[r] <- TRUE
  }

  if (r %% 50 == 0) message(sprintf("  Replicate %d / %d", r, n_rep))
}

# Compute RMSE for converged replicates
rmse <- function(est, truth) {
  valid <- !is.na(est)
  sqrt(mean((est[valid] - truth)^2))
}

cat("\n=== Monte Carlo RMSE (n = 300, n_rep = 200) ===\n\n")
param_names <- c("shape_1", "scale_1", "shape_2", "scale_2")
for (p in seq_along(theta_true)) {
  r0 <- rmse(est_s0[, p], theta_true[p])
  r1 <- rmse(est_s1[, p], theta_true[p])
  cat(sprintf("  %-8s  Scheme 0: %.4f   Scheme 1: %.4f   ratio: %.1fx\n",
              param_names[p], r0, r1, r0 / r1))
}
cat(sprintf("\n  Convergence: Scheme 0 = %d/%d, Scheme 1 = %d/%d\n",
            sum(conv_s0), n_rep, sum(conv_s1), n_rep))
```

```{r mc-results-precomputed, eval = !run_long}
# Pre-computed results from a 200-replicate Monte Carlo run:
cat("=== Monte Carlo RMSE (n = 300, n_rep = 200, pre-computed) ===\n\n")
cat("  Parameter  Scheme 0    Scheme 1    Ratio (S0/S1)\n")
cat("  --------   --------    --------    -------------\n")
cat("  shape_1    0.6121      0.0291      21.0x\n")
cat("  scale_1    0.8434      0.0847       10.0x\n")
cat("  shape_2    0.0893      0.0409       2.2x\n")
cat("  scale_2    0.1567      0.0721       2.2x\n")
cat("\n  The fast component (shape_1) sees a 21x RMSE improvement.\n")
cat("  The slow component improves ~2x --- already well-estimated under Scheme 0.\n")
```

The pattern is clear: **periodic inspection disproportionately benefits
the worst-estimated components**. The fast component's shape RMSE drops
by a factor of roughly 21, while the slow component (already
well-estimated) sees a modest 2x improvement.


## 5. Sensitivity to inspection interval width

A natural concern is whether the benefit of Scheme 1 depends on having
a fine inspection grid. It does not. The improvement is remarkably
insensitive to the interval width $\delta$.

```{r delta-sensitivity, eval = run_long}
# Sensitivity analysis: delta in {0.1, 0.5, 1.0, 2.0}
set.seed(2024)
deltas <- c(0.1, 0.5, 1.0, 2.0)
n_rep_sens <- 100
n <- 300

rmse_by_delta <- matrix(NA, nrow = length(deltas), ncol = 4)
rownames(rmse_by_delta) <- paste0("delta=", deltas)
colnames(rmse_by_delta) <- c("shape_1", "scale_1", "shape_2", "scale_2")

for (d_idx in seq_along(deltas)) {
  delta_d <- deltas[d_idx]
  est_d <- matrix(NA, nrow = n_rep_sens, ncol = 4)

  for (r in seq_len(n_rep_sens)) {
    comp_times <- matrix(0, nrow = n, ncol = m)
    for (j in seq_len(m)) {
      comp_times[, j] <- rweibull(n, shape = alpha_true[j], scale = beta_true[j])
    }
    sys_times <- apply(comp_times, 1, max)

    df_d <- data.frame(t = sys_times)
    for (j in seq_len(m)) {
      df_d[[paste0("comp_lower_", j)]] <- floor(comp_times[, j] / delta_d) * delta_d
      df_d[[paste0("comp_upper_", j)]] <- floor(comp_times[, j] / delta_d) * delta_d + delta_d
    }

    res_d <- tryCatch(solver_s1(df_d, n_starts = 3L),
                      error = function(e) NULL)
    if (!is.null(res_d) && res_d$converged) {
      est_d[r, ] <- res_d$par
    }
  }

  for (p in 1:4) {
    rmse_by_delta[d_idx, p] <- rmse(est_d[, p], theta_true[p])
  }
  message(sprintf("  delta = %.1f complete", delta_d))
}

cat("\n=== RMSE by inspection interval width ===\n\n")
print(round(rmse_by_delta, 4))
```

```{r delta-precomputed, eval = !run_long}
cat("=== RMSE by inspection interval width (pre-computed) ===\n\n")
cat("  delta    shape_1   scale_1   shape_2   scale_2\n")
cat("  -----    -------   -------   -------   -------\n")
cat("  0.1      0.0195    0.0583    0.0380    0.0685\n")
cat("  0.5      0.0291    0.0847    0.0409    0.0721\n")
cat("  1.0      0.0437    0.1102    0.0425    0.0748\n")
cat("  2.0      0.0812    0.1654    0.0461    0.0789\n")
cat("\n")
cat("  Even delta = 2.0 (coarser than the median component lifetime)\n")
cat("  gives shape_1 RMSE of 0.08 vs 0.61 under Scheme 0: a 7.5x improvement.\n")
```

The key finding: **even very coarse inspection intervals provide
substantial improvement**. Going from $\delta = 0.1$ to $\delta = 2.0$
degrades the fast component's RMSE by only a factor of 4, while the
gap between Scheme 0 and any Scheme 1 variant remains an order of
magnitude.


## 6. Information-theoretic interpretation

Under Scheme 0, the fast component's information comes through
$f_1(t) \cdot \prod_{k \neq 1} F_k(t)$, but $F_1(T) \approx 1$ makes
this nearly constant with respect to Component 1's parameters --- a flat
likelihood surface. Under Scheme 1, the interval-censoring term
$F_j(a_j^+) - F_j(a_j^-)$ varies meaningfully with the parameters even
for the fast component: two inspection intervals suffice to distinguish
increasing from decreasing hazard.

```{r information-anatomy}
# Demonstrate: how much does F_j(a+) - F_j(a-) vary with shape?
t_fail <- 1.0  # a typical failure time for the fast component
delta <- 0.5
a_lower <- floor(t_fail / delta) * delta
a_upper <- a_lower + delta

shapes_grid <- seq(0.5, 3.0, by = 0.1)
probs <- sapply(shapes_grid, function(alpha) {
  pweibull(a_upper, shape = alpha, scale = 2.0) -
    pweibull(a_lower, shape = alpha, scale = 2.0)
})

plot(shapes_grid, probs, type = "l", lwd = 2,
     xlab = expression(alpha[1] ~ "(shape parameter)"),
     ylab = expression(F[1](a^"+") - F[1](a^"-")),
     main = "Interval probability varies with shape",
     sub = sprintf("Interval [%.1f, %.1f), scale = 2.0", a_lower, a_upper))
abline(v = 1.5, lty = 2, col = "red")
text(1.5, max(probs) * 0.9, expression(alpha[1] == 1.5 ~ "(truth)"),
     pos = 4, col = "red")
```

The interval probability has substantial curvature as a function of
the shape parameter. This is precisely the information that Scheme 0
lacks --- and that Scheme 1 provides for free.


## 7. Fisher information comparison

The `compare_fisher_info()` function quantifies the total information
content of each scheme using the determinant of the observed Fisher
information matrix. The determinant ratio
$\det(I_{\text{Scheme } k}) / \det(I_{\text{Scheme } \ell})$
is a scalar summary of relative efficiency, with values less than 1
indicating that Scheme $\ell$ is more informative.

```{r fisher-demo, eval = run_long}
fi <- compare_fisher_info(
  shapes = c(1.5, 2.0),
  scales = c(2.0, 3.0),
  n = 200, delta = 0.5, n_rep = 30,
  family = "weibull"
)

cat("=== Fisher Information Determinant Comparison ===\n\n")
cat(sprintf("  Median det(I), Scheme 0: %.4e\n", fi$median_det["scheme0"]))
cat(sprintf("  Median det(I), Scheme 1: %.4e\n", fi$median_det["scheme1"]))
cat(sprintf("  Median det(I), Scheme 2: %.4e\n", fi$median_det["scheme2"]))
cat(sprintf("\n  Efficiency ratios:\n"))
cat(sprintf("    Scheme 0 / Scheme 1 = %.4f  (Scheme 1 is %.1fx more informative)\n",
            fi$efficiency_01, 1 / fi$efficiency_01))
cat(sprintf("    Scheme 1 / Scheme 2 = %.4f  (Scheme 1 recovers %.0f%% of complete-data info)\n",
            fi$efficiency_12, fi$efficiency_12 * 100))
cat(sprintf("    Scheme 0 / Scheme 2 = %.6f\n", fi$efficiency_02))
cat(sprintf("\n  Valid replicates: Scheme 0 = %d, Scheme 1 = %d, Scheme 2 = %d\n",
            fi$n_valid["scheme0"], fi$n_valid["scheme1"], fi$n_valid["scheme2"]))
```

```{r fisher-precomputed, eval = !run_long}
cat("=== Fisher Information Determinant Comparison (pre-computed) ===\n\n")
cat("  Median det(I), Scheme 0: 3.21e+04\n")
cat("  Median det(I), Scheme 1: 1.89e+07\n")
cat("  Median det(I), Scheme 2: 3.15e+07\n")
cat("\n  Efficiency ratios:\n")
cat("    Scheme 0 / Scheme 1 = 0.0017  (Scheme 1 is ~590x more informative)\n")
cat("    Scheme 1 / Scheme 2 = 0.60    (Scheme 1 recovers ~60% of complete-data info)\n")
cat("    Scheme 0 / Scheme 2 = 0.0010\n")
cat("\n  Scheme 1 with delta = 0.5 closes most of the gap between black-box\n")
cat("  observation and complete monitoring, using only periodic access.\n")
```

Scheme 0 retains only ~0.1% of complete-data information (dominated by
near-total loss for the fast component), while Scheme 1 with
$\delta = 0.5$ recovers ~60% --- a three-order-of-magnitude improvement.
The remaining 40% gap reflects the intrinsic cost of interval censoring.


## 8. Practical guidelines

The results in this vignette point to clear practical recommendations:

**When to use periodic inspection.** Whenever component-level parameter
estimates are needed and continuous monitoring is infeasible. The
information gain from even coarse inspection is dramatic --- particularly
for components whose lifetimes are short relative to the system.

**Choosing the inspection interval.** The choice of $\delta$ is
forgiving. A reasonable rule of thumb: set $\delta$ to be roughly the
median component lifetime. Even at this coarse resolution, the RMSE
improvement over Scheme 0 is typically an order of magnitude for the
worst-estimated parameters. Finer inspection helps, but with diminishing
returns.

**When Scheme 0 suffices.** If all components have similar lifetime
distributions (homogeneous or near-homogeneous system), the information
asymmetry is mild, and Scheme 0 may provide adequate estimates. The
problem is acute when component lifetimes span a wide range.

**The cost-benefit tradeoff.** Periodic inspection has real operational
cost, but a single inspection during the system's lifetime is better
than none, and a few inspections capture most of the available
information. The case for periodic inspection is strong: it transforms
an ill-conditioned estimation problem into a well-conditioned one at
modest cost.
