---
title: "Breaking Permutation Symmetry in k-out-of-n Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Breaking Permutation Symmetry in k-out-of-n Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(kofn)
options(digits = 3)
```

# The symmetry problem

The system density $f_{\text{sys}}(t; \boldsymbol{\lambda})$ for a
k-out-of-n system is invariant under permutation of the component
parameters.  This creates $m!$ equivalent modes on the likelihood
surface.  For $m = 4$, that is 24 modes --- the optimizer typically
converges to the symmetric point where all rates are equal.

The binary identification threshold demonstrated below is a statement
about information in the *likelihood*, the object that
`likelihood.model` formalizes as a generic across all reliability
packages in the rlang stack.  Each of the six mechanisms modifies the
likelihood function (by adding new observation-level terms) rather
than the estimator.  That is why every mechanism can be implemented
without leaving the ecosystem interface: define a new negative
log-likelihood, pass it to `solve_mle`, and inherit the full
`algebraic.mle` inference machinery on the result.  See
`vignette("ecosystem")` for the map.

We compare **six mechanisms** for breaking this symmetry, all on the
same baseline: $m = 4$ exponential components with rates
$\boldsymbol{\lambda} = (0.4, 0.6, 0.8, 1.0)$, $n = 300$
observations from a 2-out-of-4 system ($k = 2$).

```{r baseline}
m <- 4
k <- 2
rates <- c(0.4, 0.6, 0.8, 1.0)
rates_sorted <- sort(rates)
n <- 300
model <- kofn(k = k, m = m, family = "exponential")
set.seed(2026)
```


# 1. Scheme 0 (baseline --- no symmetry breaking)

```{r scheme0}
dat0 <- rdata_system(model$system, par = rates, n = n)
res0 <- fit_system(dat0$t, model$system, n_starts = 5)
est0 <- sort(coef(res0))
err0 <- mean(abs(est0 - rates_sorted))
cat("Estimates:", round(est0, 3), "\n")
cat("Mean error:", round(err0, 3), "\n")
```


# 2. Ordering constraints

**Idea:** If we know the component ordering (e.g., by age or
design), we can constrain $\lambda_1 \le \lambda_2 \le \cdots \le
\lambda_m$.  This reduces the $m!$ modes to a single one.

**Plausibility:** High.  Component age, material grade, or
manufacturer often provides a known reliability ordering.

**Implementation:** Reparameterize as cumulative increments:
$\lambda_j = \delta_1 + \cdots + \delta_j$ where $\delta_j > 0$.
Optimize over unconstrained $\delta$.

```{r ordering}
ll_sys <- function(par) loglik_system(dat0$t, model$system, par)
neg_ll_ordered <- function(delta) {
  if (any(delta <= 0)) return(.Machine$double.xmax / 2)
  lam <- cumsum(delta)
  val <- -ll_sys(lam)
  if (!is.finite(val)) return(.Machine$double.xmax / 2)
  val
}

# Initialize: equal increments
delta0 <- diff(c(0, rep(mean(rates), m)))
res_ord <- stats::optim(delta0, neg_ll_ordered, method = "L-BFGS-B",
                        lower = rep(1e-10, m))
est_ord <- cumsum(res_ord$par)
err_ord <- mean(abs(est_ord - rates_sorted))
cat("Estimates:", round(est_ord, 3), "\n")
cat("Mean error:", round(err_ord, 3), "\n")
```


# 3. Heterogeneous system structures

**Idea:** Observe the *same* component types deployed in systems
with *different* k values.  The different censoring patterns provide
complementary information that breaks symmetry.

**Plausibility:** Moderate.  In fleet management, the same component
type may be used in different system configurations (e.g., some
aircraft have dual-redundant, others triple-redundant hydraulics).

**Implementation:** Pool log-likelihoods from multiple k values.

```{r heterogeneous}
# Generate data from two different k configurations
set.seed(2026)
sys_k2 <- kofn_system(2, m)
sys_k3 <- kofn_system(3, m)
dat_k2 <- rdata_system(sys_k2, par = rates, n = n / 2)
dat_k3 <- rdata_system(sys_k3, par = rates, n = n / 2)

neg_ll_hetero <- function(par) {
  if (any(par <= 0)) return(.Machine$double.xmax / 2)
  ll2 <- loglik_system(dat_k2$t, sys_k2, par)
  ll3 <- loglik_system(dat_k3$t, sys_k3, par)
  val <- -(ll2 + ll3)
  if (!is.finite(val)) return(.Machine$double.xmax / 2)
  val
}

res_het <- solve_mle(neg_ll_hetero, par0 = rep(0.5, m),
                          n_par = m, n_starts = 5, nobs = n)
est_het <- sort(coef(res_het))
err_het <- mean(abs(est_het - rates_sorted))
cat("Estimates:", round(est_het, 3), "\n")
cat("Mean error:", round(err_het, 3), "\n")
```


# 4. Partial autopsy

**Idea:** After system failure, inspect $r$ of the $m$ components
to determine whether each inspected component failed or survived.
Uninspected components have unknown status.

**Plausibility:** Very high.  Post-failure inspection of accessible
components is standard practice.  Some components may be physically
inaccessible (inside a sealed unit) while others can be visually
checked.

**Implementation:** Observe a random subset of components; for each,
learn failed/survived.  Marginalize over unknown components.

```{r partial-autopsy}
# Generate full component data
set.seed(2026)
comp_times <- matrix(0, nrow = n, ncol = m)
for (j in seq_len(m)) comp_times[, j] <- stats::rexp(n, rate = rates[j])
sys_times <- vapply(seq_len(n), function(i)
  system_lifetime(model$system, comp_times[i, ]), numeric(1))

# For each observation, inspect r=2 random components
r_inspect <- 2
inspected <- matrix(FALSE, nrow = n, ncol = m)
known_failed <- matrix(FALSE, nrow = n, ncol = m)
for (i in seq_len(n)) {
  idx <- sample.int(m, r_inspect)
  inspected[i, idx] <- TRUE
  known_failed[i, idx] <- comp_times[i, idx] <= sys_times[i]
}

# Autopsy likelihood: for each observation, marginalize over
# unknown component states
neg_ll_autopsy <- function(par) {
  if (any(par <= 0)) return(.Machine$double.xmax / 2)
  ll <- 0
  for (i in seq_len(n)) {
    ti <- sys_times[i]
    f_vals <- stats::dexp(ti, rate = par)
    F_vals <- stats::pexp(ti, rate = par)
    S_vals <- 1 - F_vals

    # Known constraints from inspection
    known_surv <- inspected[i, ] & !known_failed[i, ]
    known_fail <- inspected[i, ] & known_failed[i, ]
    unknown <- !inspected[i, ]

    # Components known to have failed must be in F
    # Components known to have survived must not be in F
    # Unknown components: marginalize over failed/survived
    # We enumerate all 2^(#unknown) possibilities for unknown components
    unk_idx <- which(unknown)
    n_unk <- length(unk_idx)

    total <- 0
    for (mask in seq_len(2^n_unk) - 1L) {
      bits <- as.logical(as.integer(intToBits(mask))[seq_len(n_unk)])

      # Build candidate failed set: known_fail + unknown[bits==TRUE]
      F_cand <- which(known_fail)
      F_cand <- c(F_cand, unk_idx[bits])

      # Must have exactly k failures
      if (length(F_cand) != k) next

      not_F <- setdiff(seq_len(m), F_cand)

      # Check consistency: known survivors must not be in F
      if (any(known_surv[F_cand])) next

      # Compute term: sum_j f_j * prod_{l in F\j} F_l * prod_{l not in F} S_l
      for (j in F_cand) {
        if (f_vals[j] <= 0) next
        others_F <- setdiff(F_cand, j)
        log_term <- log(f_vals[j])
        if (length(others_F) > 0) {
          if (any(F_vals[others_F] <= 0)) next
          log_term <- log_term + sum(log(F_vals[others_F]))
        }
        if (length(not_F) > 0) {
          if (any(S_vals[not_F] <= 0)) next
          log_term <- log_term + sum(log(S_vals[not_F]))
        }
        total <- total + exp(log_term)
      }
    }
    if (total <= 0) return(.Machine$double.xmax / 2)
    ll <- ll + log(total)
  }
  -ll
}

res_aut <- solve_mle(neg_ll_autopsy, par0 = rep(0.5, m),
                          n_par = m, n_starts = 5, nobs = n)
est_aut <- sort(coef(res_aut))
err_aut <- mean(abs(est_aut - rates_sorted))
cat("Inspected:", r_inspect, "of", m, "components per observation\n")
cat("Estimates:", round(est_aut, 3), "\n")
cat("Mean error:", round(err_aut, 3), "\n")
```


# 5. Periodic inspection (Scheme 1)

**Plausibility:** High for systems that can be periodically accessed
but not continuously monitored (industrial equipment, fleet vehicles).

```{r scheme1}
set.seed(2026)
s1gen <- rdata_scheme1(model)
dat1 <- s1gen(theta = rates, n = n, delta = 0.5)
fitter1 <- fit_scheme1(model)
res1 <- fitter1(dat1, n_starts = 5)
est1 <- sort(coef(res1))
err1 <- mean(abs(est1 - rates_sorted))
cat("Estimates:", round(est1, 3), "\n")
cat("Mean error:", round(err1, 3), "\n")
```


# 6. Masked failed sets (exact)

**Plausibility:** Moderate.  Requires post-failure diagnostics that
identify which components failed, but not their exact failure times.

```{r masked}
set.seed(2026)
gen_m <- rdata_masked(model)
df_m <- gen_m(theta = rates, n = n, p_mask = 0)

ll_m <- loglik_masked(model)
neg_ll_m <- function(p) {
  val <- -ll_m(df_m, p)
  if (!is.finite(val)) return(.Machine$double.xmax / 2)
  val
}
res_mask <- solve_mle(neg_ll_m, par0 = rep(0.5, m),
                           n_par = m, n_starts = 5, nobs = n)
est_mask <- sort(coef(res_mask))
err_mask <- mean(abs(est_mask - rates_sorted))
cat("Estimates:", round(est_mask, 3), "\n")
cat("Mean error:", round(err_mask, 3), "\n")
```


# Comparison

```{r comparison}
results <- data.frame(
  mechanism = c("Scheme 0 (baseline)", "Ordering constraints",
                "Heterogeneous k", "Partial autopsy (r=2)",
                "Periodic inspection", "Masked failed sets"),
  mean_err = c(err0, err_ord, err_het, err_aut, err1, err_mask),
  plausibility = c("Always available", "High (domain knowledge)",
                    "Moderate (fleet data)", "Very high (standard practice)",
                    "High (periodic access)", "Moderate (diagnostics)")
)
results$improvement <- sprintf("%.0fx",
  results$mean_err[1] / results$mean_err)
results <- results[order(results$mean_err), ]
rownames(results) <- NULL
knitr::kable(results, digits = 3,
             caption = "Symmetry-breaking mechanisms for 2-of-4 exponential system")
```


# Discussion

The mechanisms fall into three categories:

**No additional data required:**

- *Ordering constraints* use domain knowledge (component age, grade)
  to eliminate equivalent modes.  This is the cheapest intervention
  and should always be applied when a reliability ordering is known.

**Post-failure information:**

- *Partial autopsy* inspects $r$ of $m$ components after failure.
  Even $r = 2$ out of $m = 4$ provides substantial improvement.
  This is the most **plausible** mechanism: checking accessible
  components after a failure is standard maintenance practice.
- *Masked failed sets* is the limiting case of autopsy where all
  components are inspected ($r = m$).  The improvement over partial
  autopsy quantifies the value of inspecting every component.

**During-operation monitoring:**

- *Periodic inspection* adds interval-censored component lifetime
  data.  This provides the most information per observation because
  it constrains the component failure times, not just their
  failed/survived status.
- *Heterogeneous structures* exploit fleet diversity --- different
  system configurations create different censoring patterns on shared
  parameters.  This requires no component-level data, only system-level
  observations from multiple configurations.

The practical recommendation depends on what data is available:

1. Always apply ordering constraints if a reliability ordering is known
2. If post-failure inspection is possible, do partial autopsy (even
   checking 2 of 4 components helps substantially)
3. If periodic access is feasible, add inspection intervals
4. If fleet data spans multiple configurations, pool across k values
```